\CRANpkg{lg}: An R package for Local Gaussian Approximations

### 2021-09-10 otneim.tex
Submission to the R Journal after accept

\citet{tjos:huft:2013} propose the \emph{local Gaussian correlation} (LGC) as a new measure of statistical dependence between two stochastic variables $X_1$ and $X_2$, which has the following important property yet unrivaled in the literature: It can separate between positive and negative nonlinear dependence while still reducing to the ordinary Pearson correlation coefficient if $X_1$ and $X_2$ are jointly normally distributed. The R-package \CRANpkg{localgauss} \citep{bere:klep:tjos:2014} provides two important functions in this context; one that calculates the sample LGC based on observed values of $\left(X_1,X_2\right)$, and one that uses the estimated LGC to perform a local test of independence between $X_1$ and $X_2$ as described in detail by \citet{bere:tjos:2014}.

We have lately seen a number of new applications of the LGC that the \CRANpkg{localgauss}-package does not support, however. \citet{stov:tjos:huft:2014} use the LGC to test for financial contagion across markets during crises. \citet{otne:tjos:2017} present a procedure for estimating multivariate density functions via the LGC, which \citet{otne:tjos:2018} modify in order to compute estimates of conditional density functions. \citet{laca:tjos:2017} present a test for serial independence within a time series, which \citet{laca:tjos:2018} extend in order to include a test for cross-correlation between two time series. Finally, \citet{otne:tjos:2019} develop the local Gaussian \emph{partial} correlation (LGPC) as a measure of conditional dependence, and a corresponding test for conditional independence.

This paper describes the \CRANpkg{lg}-package \citep{otne:2019}, which provides a unified framework to implement all these methods, as well as a tool for visualizing the LGC and LGPC as dependence maps. \citet{jord:tjos:2017} use the LGC in spectral analysis of time series, but those methods have their own computational ecosystem in the \pkg{localgaussSpec}-package, \citet{jord:2018}.

@@

A different approach is to estimate $f_X\left(\cdot\right)$ without any prior parametric assumptions. The classical method for non-parametric density estimation is the kernel estimator 
$$\widehat f_X\left(x\right) = \frac{1}{nb}\sum_{i=1}^n K\left(\frac{X_i - x}{b}\right),$$
where $K$ is a symmetric density function (the kernel) and $b$ is a tuning parameter (the bandwidth) that controls the smoothness of the estimate $\widehat f_X\left(\cdot\right)$. See \citet{silv:1986} for an introduction to this topic. There is also a massive statistical literature on density estimation containing extensions and improvements to the classical methods to be used in various practical situations.

\citet{hjor:jone:1996} provide one such idea. They consider a parametric family $F_{\theta}$, but instead of searching for a single parameter value $\theta_0$, for which $f_X\left(x\right) = f\left(x;\theta_0\right)$ (or approximately so), they rather assert that different members of $F_{\theta}$ may approximate $f_X$ locally in different parts of its domain. In other words, they seek to estimate a parameter \emph{function} $\theta_0\left(x\right)$ for which $f_X\left(x\right) = f\left(x;\theta_0\left(x\right)\right)$ (or approximately so), and do this by maximizing a \emph{local likelihood function} in each point $x$:
\begin{align}
\widehat\theta\left(x\right) 
&= \arg\max_{\theta \in \Theta} L_n\left(\theta, x\right) \nonumber\\ 
&= \arg\max_{\theta \in \Theta} \frac{1}{nb}\sum_{i=1}^n K\left(\frac{X_i - x}{b}\right)\log f\left(X_i; \theta\right) - \int \frac{1}{b}K\left(\frac{y - x}{b}\right)f\left(y;\theta\right)\,\textrm{d}y,
\label{eq:loclik}
\end{align}
where, again, $K$ is a symmetric density function and $b$ is a bandwidth parameter that controls the smoothness of the estimate. The second term in the local likelihood function is a penalty that ensures that the estimated density $\widehat f_X\left(x\right) = f\left(x; \widehat\theta\left(x\right)\right)$ converges correctly to the true density function $f_X\left(x\right)$ as the sample size $n$ increases to infinity, and the bandwidth $b$ decreases towards zero. See \citet{hjor:jone:1996} for a detailed discussion about this construction.

\citet{tjos:huft:2013} consider the bivariate case $X = \left(X_1, X_2\right)$ and take $F_{\theta}$ to be the family of bivariate normal distributions consisting of densities on the form

@@

In Figure \ref{fig:initial-example} we have plotted the estimated LGC for two bivariate data sets on a grid; 1000 simulated observations from a binormal distribution having correlation equal to 0.5, and the daily return on the CAC40 and FTSE100 stock indices on 1000 consecutive trading days starting on May 5th 2014 \citep{datastream}. In the first panel, we see that the estimated local correlation coincides with the global correlation, except for the estimation error which is comparable to the uncertainty observed in other non-parametric estimation methods such as the kernel density estimator (see, for instance, \citet{otne:tjos:2017} for a formal asymptotic analysis of relevant convergence rates). In the second panel we see clearly that the local correlation, and thus the dependence, is stronger in the lower left and upper right regions of the distribution than in the central parts. The phenomenon of local dependence is well known in the financial literature, and using the LGC it can be measured, interpreted, and visualized in a natural way. The interpretation of this particular figure is that extreme observations on the two stock indices are more strongly dependent than the less extreme observations.

@@

Taking the LGC as a measure of dependence opens up for a number of possibilities to construct statistical tests. \citet{bere:tjos:2014} show that $\rho\left(x\right) \equiv 0$ implies that $X_1$ and $X_2$ are independent. They show further that independence between $X_1$ and $X_2$ implies $\rho\left(x\right) \equiv 0$ if the population values of the local mean and standard deviation functions satisfy the following conditions: $\mu_i\left(x_1, x_2\right) = \mu_i\left(x_i\right)$ and $\sigma_i\left(x_1, x_2\right) = \sigma_i\left(x_i\right)$ for $i=1,2$. Equivalence between independence and $\rho\left(x\right) \equiv 0$ holds in general if the observations have been suitably transformed according to a procedure presented later in this section. It follows then that departures from $\widehat\rho\left(x\right) \equiv 0$ may be taken as evidence against the hypothesis that $X_1$ and $X_2$ are statistically independent. \citet{bere:tjos:2014} formalize this notion by testing whether $\rho\left(x\right) \equiv 0$ for all $x\in S \subset \mathbb{R}^2$ using the test statistic

@@

Consider next the stationary time series $\left\{X_t\right\}$. The autocorrelation function (ACF) $\rho_k = \rho\left(X_t, X_{t-k}\right)$ is a well known concept for describing the serial dependence in the time series, but the ACF is, again, only capable to completely capture \emph{linear} serial dependence. \citet{laca:tjos:2017} seek to remedy this by rather calculating the the local correlation between $X_t$ and $X_{t-k}$. This leads to a test for serial independence in the natural way. In fact, this work is mainly a theoretical exercise in order to accommodate time series dependence. Testing for independence between $X_t$ and $X_{t-k}$ using observations $\left\{X_t, X_{t-k}\right\}_{t = k+1}^T$ leads to the same test statistic \eqref{eq:test-statistic1} and bootstrap procedure as the test for independence between $X_1$ and $X_2$ that we described above.

\citet{laca:tjos:2018} extend this problem to test for serial cross-dependence between two time series $\left\{X_t, Y_t\right\}$ by measuring the LGC between $X_t$ and $Y_{t-k}$. Departures from $\widehat\rho\left(x,y\right) \equiv 0$ is again taken as evidence against independence, and the test statistic (\ref{eq:test-statistic1}) provides an aggregate measure of this discrepancy in the specified region $S$. In this case, however, we can not obtain replicates of the test statistic under the null hypothesis by simple permutations of the data. \citet{laca:tjos:2018} suggest rather two block bootstrap procedures to this end, using fixed and random block sizes respectively. The tests for serial dependence, and serial cross-dependence are both implemented in the \CRANpkg{lg}-package, as we demonstrate in Section \ref{chap:indtests}.

We find another application of the local Gaussian approximation in the work by \citet{stov:tjos:huft:2014}, who measure and test for financial contagion. They define contagion by "a significant increase in cross-market linkages after a shock to one country" \citep[p. 2223]{forb:rigo:2002} and employ the LGC to quantify this potential linkage. The authors estimate the LGC on a grid $\left\{x_1, x_2\right\}_k$, $k=1,\ldots,K$ along the diagonal $D=\left\{\left(x_1,x_2\right): x_1 = x_2\right\}$ \emph{before} and \emph{after} some critical event in the financial markets, denoted as the crisis (C) and the non-crisis (NC) periods respectively. They compare the two estimates using the following test statistic; 
$$T_{n,b}^D = \sum_{k=1}^K \left\{\widehat\rho_C\left(x_k, x_k\right) - \widehat\rho_{NC}\left(x_k, x_k\right) \right\}w\left(x_k,x_k\right),$$

where $w\left(\cdot, \cdot\right)$ is a weight function that serve the same purpose as the integration area $S$ in (\ref{eq:test-statistic1}). In this case, \citet{stov:tjos:huft:2014} show that a standard bootstrap will suffice in order to produce approximate replicates of $T_{n,b}^D$ under the null hypothesis of no financial contagion, and we demonstrate the implementation of this test using the \CRANpkg{lg}-package in Section \ref{chap:cont}.

Although the original work by \citet{hjor:jone:1996} provide a general framework for local likelihood density estimation using any $p$-variate parametric family as the local family, it is evident that the method may struggle in multivariate applications much in the same way as the kernel density estimator does. This is a consequence of the curse of dimensionality, the effect of which is sought remedied by an algorithm provided by \citet{otne:tjos:2017}. The idea is to fit the $p$-variate normal distribution $\psi\left(\mu, \Sigma\right)$ locally, where $\mu$ is the vector of $p$ expectations, and $\Sigma$ is the $p\times p$ covariance matrix (to which the correlation matrix $R$ corresponds), but under the following structural simplifications:

\begin{align}
\mu_i\left(x\right) &= \mu_i\left(x_1, \ldots, x_p\right) \stackrel{\textrm{def}}{=} \mu_i\left(x_i\right) \label{eq:simp-mu}\\
\sigma_i\left(x\right) &= \sigma_i\left(x_1, \ldots, x_p\right)  \stackrel{\textrm{def}}{=} \sigma_i\left(x_i\right) \label{eq:simp-sig} \\
\rho_{ij}\left(x\right) &= \rho_{ij}\left(x_1, \ldots, x_p\right) \stackrel{\textrm{def}}{=} \rho_{ij}\left(x_i, x_j\right). \label{eq:simp-rho}
\end{align}
\citet{otne:tjos:2017} estimate the local parameters above by first obtaining univariate marginal locally Gaussian fits (eqs. \ref{eq:simp-mu} and \ref{eq:simp-sig}), and then pairwise bivariate locally Gaussian fits (eq. \ref{eq:simp-rho}). In the second step the estimates $\widehat\mu_i\left(x_i\right)$, $\widehat\mu_j\left(x_j\right)$, $\widehat\sigma_i\left(x_i\right)$ and $\widehat\sigma_j\left(x_j\right)$ are kept fixed in the estimation of the pairwise local correlation. They argue further that the following transformation technique will produce better density estimates in many situations. The motivation for introducing the simplifications defined in equations \ref{eq:simp-mu}-\ref{eq:simp-rho} can be compared to the practical advantages of estimating additive regression models, where $\textrm{E}\left(Y\right) = f\left(x_1, \ldots, x_p\right) \stackrel{\textrm{def}}{=} f_1\left(x_1\right) + \cdots + f_p\left(x_p\right)$.

@@

\begin{equation}
\mu_i\left(z_i\right) \stackrel{\textrm{def}}{=} 0 \,\, \textrm{ and } \sigma_i\left(z_i\right) \stackrel{\textrm{def}}{=} 1, \,\, i = 1,\ldots,p,
\label{eq:simp}
\end{equation}
so that the only parameter functions left to estimate are the pairwise local Gaussian correlations $R\left(z\right) = \left\{\rho_{ij}\left(z_i, z_j\right)\right\}_{i<j}$. We use the notation $Z$, $z_i$ and $z_j$ to signify that the estimation is performed on the (approximate) standard normal scale, or $z$-scale for short. We can then estimate the joint density $f_Z\left(z\right)$ of $Z$ as
\begin{equation}
\widehat f_Z\left(z\right) = \psi\left(z; \mu\left(z\right) = 0, \sigma\left(z\right) = 1, R = \widehat R\left(z\right)\right),
\label{eq:transformed-density}
\end{equation}
where $\mu\left(z\right) = \left\{\mu_i\left(z\right)\right\}$ and $\sigma\left(z\right) = \left\{\sigma_i\left(z\right)\right\}$ for $i=1,\ldots,p$, and then substitute $f_Z$ for $\widehat f_Z$ in the following relation obtained by \citet{otne:tjos:2017} in order to estimate the unknown density $f_X$:
\begin{equation}
f\left(x\right) = f_{Z}\big(\Phi^{-1}\left(F_1\left(x_1\right)\right), \ldots, \Phi^{-1}\left(F_p\left(x_p\right)\right)\big) \times \prod_{i=1}^p \frac{f_i\left(x_i\right)}{\phi\big(\Phi^{-1}\left(F_i\left(x_i\right)\right)\big)},
\label{eq:backtrans}
\end{equation}
where $\phi\left(\cdot\right)$ is the standard normal pdf. This estimator is implemented the \CRANpkg{lg}-package as demonstrated in Section \ref{chap:density}.

@@

\begin{align}
\mu^* &= \mu_1 + \Sigma_{12}\Sigma_{22}^{-1}\left(x^{\left(2\right)} - \mu_2\right) \label{eq:cond-mu} \\
\Sigma^* &= \Sigma_{11} - \Sigma_{12}\Sigma_{22}^{-1}\Sigma_{21}, \label{eq:cond-sig}
\end{align}
see e.g. \citet[chapter 4]{john:wich:2007}. \citet{otne:tjos:2018} demonstrate that this property may be translated into a corresponding local argument without modification. That is, if the joint density $f_X\left(\cdot\right)$ can be written on a locally Gaussian form 
$$f_X\left(x\right) = \psi\left(x, \mu\left(x\right), \Sigma\left(x\right)\right),$$
then the conditional density of $X^{\left(1\right)}|X^{\left(2\right)} = x^{\left(2\right)}$ can also be written on the same locally Gaussian form with local parameters given by equations (\ref{eq:cond-mu}) and (\ref{eq:cond-sig}), except that all quantities are $x$-dependent. If we use the transformation technique described above together with simplification (\ref{eq:simp}), the local versions of equations (\ref{eq:cond-mu}) and (\ref{eq:cond-sig}) simplify to

@@

Finally, we refer to \citet{otne:tjos:2019} who take the local version of the conditional covariance matrix (\ref{eq:cond-sig}) (or (\ref{eq:cond-sig-trans}) in the transformed case) as a measure for conditional dependence, and thus as an instrument to test for conditional independence. Consider the stochastic vector $X = \left(X^{\left(1\right)}, X^{\left(2\right)}, X^{\left(3\right)}\right)$, where $X^{\left(1\right)}$ and $X^{\left(2\right)}$ are scalars and $X^{\left(3\right)}$ may be a vector. $X^{\left(1\right)}$ is conditionally independent from $X^{\left(2\right)}$ given $X^{\left(3\right)}$, written $X^{\left(1\right)} \perp X^{\left(2\right)} \,|\,X^{\left(3\right)}$, if the stochastic variables $X^{\left(1\right)}\,|\,X^{\left(3\right)}$ and $X^{\left(2\right)}\,|\,X^{\left(3\right)}$ are independent, or, equivalently, if the joint conditional density of $X^{\left(1\right)}$ and $X^{\left(2\right)}$ given $X^{\left(3\right)}$ can be written as the product

\begin{equation}
f_{X^{\left(1\right)},X^{\left(2\right)}|X^{\left(3\right)}}\left(x^{\left(1\right)}, x^{\left(2\right)}|x^{\left(3\right)}\right) = f_{X^{\left(1\right)}|X^{\left(3\right)}}\left(x^{\left(1\right)}|x^{\left(3\right)}\right) \times f_{X^{\left(2\right)}|X^{\left(3\right)}}\left(x^{\left(2\right)}|x^{\left(3\right)}\right).
\label{eq:cond-ind}
\end{equation}
In this case, denote by $\alpha\left(z\right)$ the off-diagonal element in the $2\times2$ local correlation matrix $R^*\left(z\right)$ (which derives directly from $\Sigma^*\left(z\right)$ as given in (\ref{eq:cond-sig-trans})). If $X$ has a local Gaussian distribution, the conditional independence (\ref{eq:cond-ind}) is equivalent to $\alpha\left(z\right) \equiv 0$, and \citet{otne:tjos:2019} take departures from this relation as evidence against the hypothesis of conditional independence between $X^{\left(1\right)}$ and $X^{\left(2\right)}$ given $X^{\left(3\right)}$. The natural way to quantify this is the test functional

\begin{equation}
T^{CI}_{n, b} = \int h\left(\widehat\alpha\left(z\right)\right) \, \textrm{d}F_n(z).
\label{eq:ci-test-statistic}
\end{equation}
\citet{otne:tjos:2019} describe a bootstrap procedure for generating replicates of $T_{n,b}^{CI}$ under the null hypothesis. In Section \ref{chap:partial} we demonstrate how the \CRANpkg{lg}-package may be used to extract estimates of the local partial correlation and perform tests for conditional independence according to this scheme.

@@

\textbf{4. A full locally Gaussian fit for trivariate data.} If the number of variables $p$ is equal to $3$ \emph{and} we choose to transform the data to marginal standard normality (see the next sub-section), the transformed density $f_Z\left(\cdot\right)$ in (\ref{eq:transformed-density}) may be estimated by jointly estimating the three local correlations $\rho_{12}\left(z_1, z_2, z_3\right)$, $\rho_{13}\left(z_1, z_2, z_3\right)$ and $\rho_{23}\left(z_1, z_2, z_3\right)$. This estimation method was introduced recently by \citet{otne:tjos:2019} in order to increase power of their conditional independence test, but it can be used in any application described in this paper that consider trivariate data. To use this estimation method, create the lg-object by running

@@

\section{Statistical inference using the \texttt{lg}-package}
\label{chap:inference}

>>>>

\cite{tjos:huft:2013} propose the \emph{local Gaussian correlation} (LGC) as a new measure of statistical dependence between two stochastic variables $X_1$ and $X_2$, which has the following important property yet unrivaled in the literature: It can separate between positive and negative nonlinear dependence while still reducing to the ordinary Pearson correlation coefficient if $X_1$ and $X_2$ are jointly normally distributed. The R-package \CRANpkg{localgauss} \citep{bere:klep:tjos:2014} provides two important functions in this context; one that calculates the sample LGC based on observed values of $\left(X_1,X_2\right)$, and one that uses the estimated LGC to perform a local test of independence between $X_1$ and $X_2$ as described in detail by \cite{bere:tjos:2014}.

We have lately seen a number of new applications of the LGC that the \CRANpkg{localgauss}-package does not support, however. \cite{stov:tjos:huft:2014} use the LGC to test for financial contagion across markets during crises. \cite{otne:tjos:2017} present a procedure for estimating multivariate density functions via the LGC, which \cite{otne:tjos:2018} modify in order to compute estimates of conditional density functions. \cite{laca:tjos:2017} present a test for serial independence within a time series, which \cite{laca:tjos:2018} extend in order to include a test for cross-correlation between two time series. Finally, \cite{otne:tjos:2019} develop the local Gaussian \emph{partial} correlation (LGPC) as a measure of conditional dependence, and a corresponding test for conditional independence.

This paper describes the \CRANpkg{lg}-package \citep{otne:2019}, which provides a unified framework to implement all these methods, as well as a tool for visualizing the LGC and LGPC as dependence maps. \cite{jord:tjos:2017} use the LGC in spectral analysis of time series, but those methods have their own computational ecosystem in the \pkg{localgaussSpec}-package, \cite{jord:2018}.

@@

A different approach is to estimate $f_X\left(\cdot\right)$ without any prior parametric assumptions. The classical method for non-parametric density estimation is the kernel estimator 
$$\widehat f_X\left(x\right) = \frac{1}{nb}\sum_{i=1}^n K\left(\frac{X_i - x}{b}\right),$$
where $K$ is a symmetric density function (the kernel) and $b$ is a tuning parameter (the bandwidth) that controls the smoothness of the estimate $\widehat f_X\left(\cdot\right)$. See \cite{silv:1986} for an introduction to this topic. There is also a massive statistical literature on density estimation containing extensions and improvements to the classical methods to be used in various practical situations.

\cite{hjor:jone:1996} provide one such idea. They consider a parametric family $F_{\theta}$, but instead of searching for a single parameter value $\theta_0$, for which $f_X\left(x\right) = f\left(x;\theta_0\right)$ (or approximately so), they rather assert that different members of $F_{\theta}$ may approximate $f_X$ locally in different parts of its domain. In other words, they seek to estimate a parameter \emph{function} $\theta_0\left(x\right)$ for which $f_X\left(x\right) = f\left(x;\theta_0\left(x\right)\right)$ (or approximately so), and do this by maximizing a \emph{local likelihood function} in each point $x$:
\begin{align}
\widehat\theta\left(x\right) 
&= \arg\max_{\theta \in \Theta} L_n\left(\theta, x\right) \nonumber\\ 
&= \arg\max_{\theta \in \Theta} \frac{1}{nb}\sum_{i=1}^n K\left(\frac{X_i - x}{b}\right)\log f\left(X_i; \theta\right) - \int \frac{1}{b}K\left(\frac{y - x}{b}\right)f\left(y;\theta\right)\,\textrm{d}y,
\label{eq:loclik}
\end{align}
where, again, $K$ is a symmetric density function and $b$ is a bandwidth parameter that controls the smoothness of the estimate. The second term in the local likelihood function is a penalty that ensures that the estimated density $\widehat f_X\left(x\right) = f\left(x; \widehat\theta\left(x\right)\right)$ converges correctly to the true density function $f_X\left(x\right)$ as the sample size $n$ increases to infinity, and the bandwidth $b$ decreases towards zero. See \cite{hjor:jone:1996} for a detailed discussion about this construction.

\cite{tjos:huft:2013} consider the bivariate case $X = \left(X_1, X_2\right)$ and take $F_{\theta}$ to be the family of bivariate normal distributions consisting of densities on the form

@@

In Figure \ref{fig:initial-example} we have plotted the estimated LGC for two bivariate data sets on a grid; 1000 simulated observations from a binormal distribution having correlation equal to 0.5, and the daily return on the CAC40 and FTSE100 stock indices on 1000 consecutive trading days starting on May 5th 2014 \citep{datastream}. In the first panel, we see that the estimated local correlation coincides with the global correlation, except for the estimation error which is comparable to the uncertainty observed in other non-parametric estimation methods such as the kernel density estimator (see, for instance, \cite{otne:tjos:2017} for a formal asymptotic analysis of relevant convergence rates). In the second panel we see clearly that the local correlation, and thus the dependence, is stronger in the lower left and upper right regions of the distribution than in the central parts. The phenomenon of local dependence is well known in the financial literature, and using the LGC it can be measured, interpreted, and visualized in a natural way. The interpretation of this particular figure is that extreme observations on the two stock indices are more strongly dependent than the less extreme observations.

@@

Taking the LGC as a measure of dependence opens up for a number of possibilities to construct statistical tests. \cite{bere:tjos:2014} show that $\rho\left(x\right) \equiv 0$ implies that $X_1$ and $X_2$ are independent. They show further that independence between $X_1$ and $X_2$ implies $\rho\left(x\right) \equiv 0$ if the population values of the local mean and standard deviation functions satisfy the following conditions: $\mu_i\left(x_1, x_2\right) = \mu_i\left(x_i\right)$ and $\sigma_i\left(x_1, x_2\right) = \sigma_i\left(x_i\right)$ for $i=1,2$. Equivalence between independence and $\rho\left(x\right) \equiv 0$ holds in general if the observations have been suitably transformed according to a procedure presented later in this section. It follows then that departures from $\widehat\rho\left(x\right) \equiv 0$ may be taken as evidence against the hypothesis that $X_1$ and $X_2$ are statistically independent. \cite{bere:tjos:2014} formalize this notion by testing whether $\rho\left(x\right) \equiv 0$ for all $x\in S \subset \mathbb{R}^2$ using the test statistic

@@

Consider next the stationary time series $\left\{X_t\right\}$. The autocorrelation function (ACF) $\rho_k = \rho\left(X_t, X_{t-k}\right)$ is a well known concept for describing the serial dependence in the time series, but the ACF is, again, only capable to completely capture \emph{linear} serial dependence. \cite{laca:tjos:2017} seek to remedy this by rather calculating the the local correlation between $X_t$ and $X_{t-k}$. This leads to a test for serial independence in the natural way. In fact, this work is mainly a theoretical exercise in order to accommodate time series dependence. Testing for independence between $X_t$ and $X_{t-k}$ using observations $\left\{X_t, X_{t-k}\right\}_{t = k+1}^T$ leads to the same test statistic \eqref{eq:test-statistic1} and bootstrap procedure as the test for independence between $X_1$ and $X_2$ that we described above.

\cite{laca:tjos:2018} extend this problem to test for serial cross-dependence between two time series $\left\{X_t, Y_t\right\}$ by measuring the LGC between $X_t$ and $Y_{t-k}$. Departures from $\widehat\rho\left(x,y\right) \equiv 0$ is again taken as evidence against independence, and the test statistic (\ref{eq:test-statistic1}) provides an aggregate measure of this discrepancy in the specified region $S$. In this case, however, we can not obtain replicates of the test statistic under the null hypothesis by simple permutations of the data. \cite{laca:tjos:2018} suggest rather two block bootstrap procedures to this end, using fixed and random block sizes respectively. The tests for serial dependence, and serial cross-dependence are both implemented in the \CRANpkg{lg}-package, as we demonstrate in Section \ref{chap:indtests}.

We find another application of the local Gaussian approximation in the work by \cite{stov:tjos:huft:2014}, who measure and test for financial contagion. They define contagion by "a significant increase in cross-market linkages after a shock to one country" \citep[p. 2223]{forb:rigo:2002} and employ the LGC to quantify this potential linkage. The authors estimate the LGC on a grid $\left\{x_1, x_2\right\}_k$, $k=1,\ldots,K$ along the diagonal $D=\left\{\left(x_1,x_2\right): x_1 = x_2\right\}$ \emph{before} and \emph{after} some critical event in the financial markets, denoted as the crisis (C) and the non-crisis (NC) periods respectively. They compare the two estimates using the following test statistic; 
$$T_{n,b}^D = \sum_{k=1}^K \left\{\widehat\rho_C\left(x_k, x_k\right) - \widehat\rho_{NC}\left(x_k, x_k\right) \right\}w\left(x_k,x_k\right),$$

where $w\left(\cdot, \cdot\right)$ is a weight function that serve the same purpose as the integration area $S$ in (\ref{eq:test-statistic1}). In this case, \cite{stov:tjos:huft:2014} show that a standard bootstrap will suffice in order to produce approximate replicates of $T_{n,b}^D$ under the null hypothesis of no financial contagion, and we demonstrate the implementation of this test using the \CRANpkg{lg}-package in Section \ref{chap:cont}.

Although the original work by \cite{hjor:jone:1996} provide a general framework for local likelihood density estimation using any $p$-variate parametric family as the local family, it is evident that the method may struggle in multivariate applications much in the same way as the kernel density estimator does. This is a consequence of the curse of dimensionality, the effect of which is sought remedied by an algorithm provided by \cite{otne:tjos:2017}. The idea is to fit the $p$-variate normal distribution $\psi\left(\mu, \Sigma\right)$ locally, where $\mu$ is the vector of $p$ expectations, and $\Sigma$ is the $p\times p$ covariance matrix (to which the correlation matrix $R$ corresponds), but under the following structural simplifications:

\begin{align}
\mu_i\left(x\right) &= \mu_i\left(x_1, \ldots, x_p\right) \stackrel{\textrm{def}}{=} \mu_i\left(x_i\right) \label{eq:simp-mu}\\
\sigma_i\left(x\right) &= \sigma_i\left(x_1, \ldots, x_p\right)  \stackrel{\textrm{def}}{=} \sigma_i\left(x_i\right) \label{eq:simp-sig} \\
\rho_{ij}\left(x\right) &= \rho_{ij}\left(x_1, \ldots, x_p\right) \stackrel{\textrm{def}}{=} \rho_{ij}\left(x_i, x_j\right). \label{eq:simp-rho}
\end{align}
\cite{otne:tjos:2017} estimate the local parameters above by first obtaining univariate marginal locally Gaussian fits (eqs. \ref{eq:simp-mu} and \ref{eq:simp-sig}), and then pairwise bivariate locally Gaussian fits (eq. \ref{eq:simp-rho}). In the second step the estimates $\widehat\mu_i\left(x_i\right)$, $\widehat\mu_j\left(x_j\right)$, $\widehat\sigma_i\left(x_i\right)$ and $\widehat\sigma_j\left(x_j\right)$ are kept fixed in the estimation of the pairwise local correlation. The motivation for introducing the simplifications defined in equations \ref{eq:simp-mu}-\ref{eq:simp-rho} can be compared to the practical advantages of estimating additive regression models, where $\textrm{E}\left(Y\right) = f\left(x_1, \ldots, x_p\right) \stackrel{\textrm{def}}{=} f_1\left(x_1\right) + \cdots + f_p\left(x_p\right)$. \cite{otne:tjos:2017} argue further that the following transformation technique will produce better density estimates in many situations.

@@

\begin{equation}
\mu_i\left(z_i\right) \stackrel{\textrm{def}}{=} 0 \,\, \textrm{ and } \sigma_i\left(z_i\right) \stackrel{\textrm{def}}{=} 1, \,\, i = 1,\ldots,p,
\label{eq:simp}
\end{equation}
so that the only parameter functions left to estimate are the pairwise local Gaussian correlations $R\left(z\right) = \left\{\rho_{ij}\left(z_i, z_j\right)\right\}_{i<j}$. We use the notation $Z$, $z_i$ and $z_j$ to signify that the estimation is performed on the (approximate) standard normal scale, or $z$-scale for short. We can then estimate the joint density $f_Z\left(z\right)$ of $Z$ as
\begin{equation}
\widehat f_Z\left(z\right) = \psi\left(z; \mu\left(z\right) = 0, \sigma\left(z\right) = 1, R = \widehat R\left(z\right)\right),
\label{eq:transformed-density}
\end{equation}
where $\mu\left(z\right) = \left\{\mu_i\left(z\right)\right\}$ and $\sigma\left(z\right) = \left\{\sigma_i\left(z\right)\right\}$ for $i=1,\ldots,p$, and then substitute $f_Z$ for $\widehat f_Z$ in the following relation obtained by \cite{otne:tjos:2017} in order to estimate the unknown density $f_X$:
\begin{equation}
f\left(x\right) = f_{Z}\big(\Phi^{-1}\left(F_1\left(x_1\right)\right), \ldots, \Phi^{-1}\left(F_p\left(x_p\right)\right)\big) \times \prod_{i=1}^p \frac{f_i\left(x_i\right)}{\phi\big(\Phi^{-1}\left(F_i\left(x_i\right)\right)\big)},
\label{eq:backtrans}
\end{equation}
where $\phi\left(\cdot\right)$ is the standard normal pdf. This estimator is implemented the \CRANpkg{lg}-package as demonstrated in Section \ref{chap:density}.

@@

\begin{align}
\mu^* &= \mu_1 + \Sigma_{12}\Sigma_{22}^{-1}\left(x^{\left(2\right)} - \mu_2\right) \label{eq:cond-mu} \\
\Sigma^* &= \Sigma_{11} - \Sigma_{12}\Sigma_{22}^{-1}\Sigma_{21}, \label{eq:cond-sig}
\end{align}
see e.g. \cite[chapter 4]{john:wich:2007}. \cite{otne:tjos:2018} demonstrate that this property may be translated into a corresponding local argument without modification. That is, if the joint density $f_X\left(\cdot\right)$ can be written on a locally Gaussian form 
$$f_X\left(x\right) = \psi\left(x, \mu\left(x\right), \Sigma\left(x\right)\right),$$
then the conditional density of $X^{\left(1\right)}|X^{\left(2\right)} = x^{\left(2\right)}$ can also be written on the same locally Gaussian form with local parameters given by equations (\ref{eq:cond-mu}) and (\ref{eq:cond-sig}), except that all quantities are $x$-dependent. If we use the transformation technique described above together with simplification (\ref{eq:simp}), the local versions of equations (\ref{eq:cond-mu}) and (\ref{eq:cond-sig}) simplify to

@@

Finally, we refer to \cite{otne:tjos:2019} who take the local version of the conditional covariance matrix (\ref{eq:cond-sig}) (or (\ref{eq:cond-sig-trans}) in the transformed case) as a measure for conditional dependence, and thus as an instrument to test for conditional independence. Consider the stochastic vector $X = \left(X^{\left(1\right)}, X^{\left(2\right)}, X^{\left(3\right)}\right)$, where $X^{\left(1\right)}$ and $X^{\left(2\right)}$ are scalars and $X^{\left(3\right)}$ may be a vector. $X^{\left(1\right)}$ is conditionally independent from $X^{\left(2\right)}$ given $X^{\left(3\right)}$, written $X^{\left(1\right)} \perp X^{\left(2\right)} \,|\,X^{\left(3\right)}$, if the stochastic variables $X^{\left(1\right)}\,|\,X^{\left(3\right)}$ and $X^{\left(2\right)}\,|\,X^{\left(3\right)}$ are independent, or, equivalently, if the joint conditional density of $X^{\left(1\right)}$ and $X^{\left(2\right)}$ given $X^{\left(3\right)}$ can be written as the product

\begin{equation}
f_{X^{\left(1\right)},X^{\left(2\right)}|X^{\left(3\right)}}\left(x^{\left(1\right)}, x^{\left(2\right)}|x^{\left(3\right)}\right) = f_{X^{\left(1\right)}|X^{\left(3\right)}}\left(x^{\left(1\right)}|x^{\left(3\right)}\right) \times f_{X^{\left(2\right)}|X^{\left(3\right)}}\left(x^{\left(2\right)}|x^{\left(3\right)}\right).
\label{eq:cond-ind}
\end{equation}
In this case, denote by $\alpha\left(z\right)$ the off-diagonal element in the $2\times2$ local correlation matrix $R^*\left(z\right)$ (which derives directly from $\Sigma^*\left(z\right)$ as given in (\ref{eq:cond-sig-trans})). If $X$ has a local Gaussian distribution, the conditional independence (\ref{eq:cond-ind}) is equivalent to $\alpha\left(z\right) \equiv 0$, and \cite{otne:tjos:2019} take departures from this relation as evidence against the hypothesis of conditional independence between $X^{\left(1\right)}$ and $X^{\left(2\right)}$ given $X^{\left(3\right)}$. The natural way to quantify this is the test functional

\begin{equation}
T^{CI}_{n, b} = \int h\left(\widehat\alpha\left(z\right)\right) \, \textrm{d}F_n(z).
\label{eq:ci-test-statistic}
\end{equation}
\cite{otne:tjos:2019} describe a bootstrap procedure for generating replicates of $T_{n,b}^{CI}$ under the null hypothesis. In Section \ref{chap:partial} we demonstrate how the \CRANpkg{lg}-package may be used to extract estimates of the local partial correlation and perform tests for conditional independence according to this scheme.

@@

\textbf{4. A full locally Gaussian fit for trivariate data.} If the number of variables $p$ is equal to $3$ \emph{and} we choose to transform the data to marginal standard normality (see the next sub-section), the transformed density $f_Z\left(\cdot\right)$ in (\ref{eq:transformed-density}) may be estimated by jointly estimating the three local correlations $\rho_{12}\left(z_1, z_2, z_3\right)$, $\rho_{13}\left(z_1, z_2, z_3\right)$ and $\rho_{23}\left(z_1, z_2, z_3\right)$. This estimation method was introduced recently by \cite{otne:tjos:2019} in order to increase the power of their conditional independence test, but it can be used in any application described in this paper that consider trivariate data. To use this estimation method, create the lg-object by running

@@

\section{Statistical inference using the \CRANpkg{lg}-package}
\label{chap:inference}
